Reconstructing contact network parameters from viral phylogenies

### 2016-07-01 doc/paper/paper.Rnw
Progress on paper changes

When an infectious disease spreads through a population, transmissions are
generally more likely to occur between certain pairs of individuals. Such pairs
must have a particular mode of contact with one another, which varies with the
mode of transmission of the disease. For airborne pathogens, physical proximity
may be sufficient, while for sexually transmitted diseases, sexual or in some
cases blood-to-blood contact is required. The population together with the set
of links between individuals along which transmission can occur is called the
contact network~\autocite{klovdahl1985social, morris1993epidemiology}. The
structure of the contact network underlying an epidemic can profoundly impact
the speed and pattern of the epidemic's expansion. Network structure can
influence the prevalence trajectory~\autocite{o2011contact, ma2013importance}
and epidemic threshold~\autocite{barthelemy2005dynamical}, in turn affecting
the estimates of quantities such as effective population
size~\autocite{goodreau2006assessing}.  From a public health perspective,
contact networks have been explored as tools for curtailing epidemic spread, by
way of interventions targeted to well-connected
nodes~\autocite{wang2015targeting}. True contact networks are a challenging
type of data to collect, requiring extensive epidemiological
investigation~\autocite{welch2011statistical, eames2015six}.

Viral sequence data, on the other hand, has become \del{relatively inexpensive
and straightforward to collect on a population level.} \add{easier to collect
as the cost of sequencing has declined. In the case of HIV, genotyping has
become part of routine clinical care in several health regions.'' Due to the
high mutation rate of RNA viruses, epidemiological processes impact the course
of viral evolution, thereby shaping the inter-host viral
phylogeny~\autocite{drummond2003measurably}. The term ``phylodynamics'' was
coined to describe this interaction, as well as the growing family of inference
methods to estimate epidemiological parameters from viral
phylogenies~\autocite{grenfell2004unifying}. These methods have revealed
diverse properties of local viral outbreaks, from basic reproductive
number~\autocite{stadler2012estimating}, to the degree of
clustering~\autocite{hughes2009molecular}, to the elevated transmission risk
during acute infection~\autocite{volz2012simple}. On the other hand, although
sophisticated methods have been developed for fitting complex population
genetic models to phylogenies~\autocite{volz2012complex,
rasmussen2014phylodynamic}, inference of structural network parameters has to
date been limited. However, it has been shown that network structure has a
tangible impact on phylogeny shape~\autocite{leventhal2012inferring,
colijn2014phylogenetic, goodreau2006assessing, robinson2013dynamics,
villandre2016assessment}, suggesting that such statistical inference might be
possible~\autocite{welch2011statistical}. \add{In the context of networks,
sequence data have the advantage of being objective, in that they are not
affected by misreporting. However, just as with survey data, it is important to
collect a representative sample from the population to perform accurate
inference~\autocite{novitsky2014impact}.}

Survey-based studies of sexual networks~\autocite{colgate1989risk,
liljeros2001web, schneeberger2004scale, latora2006network, rothenberg2007large,
clemenccon2015statistical} have found that these networks tend to have a degree
distribution which follows a power law \autocite[although there has been some
disagreement, see][]{handcock2004likelihood}. That is, the number of nodes of
degree $k$ is proportional to $k^{-\gamma}$ for some constant $\gamma$. These
networks are also referred to as
``scale-free''~\autocite{barabasi1999emergence}. One process by which
scale-free networks can be generated is preferential attachment, where nodes
with a high number of contacts attract new connections at an elevated rate. The
first contact network model incorporating preferential attachment was
introduced by \textcite{barabasi1999emergence}, and is now referred to as the
Barab\'asi-Albert (BA) model. Under this model, networks are formed by
iteratively adding nodes with $m$ new edges each. In the most commonly studied
formulation, these new edges are joined to existing nodes of degree $k$ with
probability proportional to $k$, so that nodes of high degree tend to attract
more connections. \citeauthor{barabasi1999emergence} suggested an extension
where the probability of attaching to a node of degree $k$ is $k^\alpha$ for
some non-negative constant $\alpha$, and we use this extension in this work.
\add{When $\alpha \neq 1$, the degree distribution is no longer a power law:
for $\alpha < 1$, the distribution is a stretched exponential, while for
$\alpha > 1$, it is a ``gelation'' type distribution where one or a few hub
nodes are connected to nearly every other node in the graph.}

@@

Standard methods of model fitting involve calculation of the likelihood of
observed data under the model. In maximum likelihood estimation, a quantity
proportional to the likelihood is optimized, often through a standard
multi-dimensional numerical optimization procedure. Bayesian methods integrate
prior information by optimizing the posterior probability instead. To avoid
calculation of a normalizing constant, Bayesian inference is often performed
using Markov chain Monte Carlo (MCMC), which uses likelihood \emph{ratios} in
which the normalizing constants cancel out. Unfortunately, it is generally
difficult to explicitly calculate the likelihood of an observed transmission
tree under a contact network model, even up to a normalizing constant. To do
so, it would be necessary to integrate over all possible networks, and also
over all possible labellings of the internal nodes of the transmission tree.
While it is not known (to us) whether such integration is tractable, a simpler
alternative is offered by likelihood-free methods, namely approximate Bayesian
computation (ABC)~\autocite{tavare1997inferring, beaumont2002approximate}. ABC
leverages the fact that, although calculating the likelihood may be
impractical, generating simulated datasets according to a model is often
straightforward. If our model fits the data well, the simulated data it
produces should be similar to the observed data. More formally, if $D$ is the
observed data, the posterior distribution $f(\theta \mid D)$ on model
parameters $\theta$ is replaced as the target of statistical inference by
$f(\theta \mid \rho(\hat{D}, D) < \varepsilon)$, where $\rho$ is a distance
function, $\hat{D}$ is a simulated dataset according to $\theta$, and
$\varepsilon$ is a small tolerance~\autocite{sunnaaker2013approximate}. \del{In
the specific case when $\rho$ is a kernel function, the approach is known as
ABC~\autocite{nakagome2013kernel, poon2015phylodynamic}.} \add{Our
group~\autocite{poon2015phylodynamic} and others~\autocite{park2015k2} have
demonstrated that taking $\rho$ to be kernel function TODO}

@@

\add{In the original formulation of ABC-SMC~\autocite{sisson2007sequential,
beaumont2009adaptive}, the user is required to specify a decreasing sequence of
tolerances $\lbrace\varepsilon_i\rbrace$. At iteration $i$, particles with no
associated simulated datasets within distance $\varepsilon_i$ of the observed
data are removed from the population. In the adaptive version of
\textcite{del2012adaptive}, the sequence of tolerances is determined
automatically by fixing the decay rate of the population's expected sample size
(ESS) to a user-defined value. \citeauthor{del2012adaptive} call this value
$\alpha$, but we will refer to it here as $\alpha_{\text{ESS}}$ to avoid
confusion with the preferential attachment power parameter of the BA model. A
computer program implementing our method is freely available at
\url{https://github.com/rmcclosk/netabc} (last accessed \today).}

@@

Nodes in our networks followed simple SI dynamics, meaning that they became
infected at a rate proportional to their number of infected neighbours, and
never recovered. For all analyses, the transmission trees' branch lengths were
scaled by dividing by their mean. We used the \textit{igraph} library's
implementation of the BA model~\autocite{csardi2006igraph} to generate the
graphs. The analyses were run on Westgrid (\url{https://www.westgrid.ca/}) and
a local computer cluster.

@@

\add{This experiment also had the secondary purpose of validating our choice of
the tree kernel as a distance measure in ABC. To tell the trees apart, we used
a classifier based on the tree kernel, but we also tested two other tree shape
statistics. Sackin's index~\autocite{shao1990tree} is a measure of tree
imbalance which not take branch lengths into account, considering only the
topology. The normalized lineages-through-time
\autocite[nLTT,][]{janzen2015approximate} compares two trees based on normalized
distributions of their branching times, and does not explicitly consider the
topology. Since the tree kernel incorporates both of these sources of
information, we expected it to outperform the other two statistics. Finally,
the tree kernel can be tuned by adjusting the values of the meta-parameters
$\lambda$ and $\sigma$ (the ``decay factor'' and ``radial basis function
variance'', see \textcite{poon2013mapping}). The results of this experiment
were used to select values for these meta-parameters to carry forward based on
their accuracy in distinguishing the different trees.}

\del{We used the phylogenetic kernel developed by \textcite{poon2013mapping}}
To test whether the parameters of the BA model had a \add{measurable} effect on
tree shape, 100 networks were simulated under each of three different values of
$\alpha$: 0.5, 1.0, and 1.5 (300 networks total). The other parameters were
fixed to the following values: $N = 5000$, $I = 1000$, and $m = 2$. A
transmission tree with 500 tips was simulated over each network (300
transmission trees total). The 300 trees were compared pairwise with the tree
kernel to form a $300 \times 300$ kernel matrix. The kernel meta-parameters
$\lambda$ \del{(the ``decay factor''),} and $\sigma$ \del{(the ``radial basis
function variance'')~\autocite[see][]{poon2013mapping},} were set to 0.3 and 4
respectively. \add{We also computed a $300 \times 300$ matrix of pairwise nLTT
values, and a $1 \times 300$ vector of Sackin's index values.} We constructed 
\add{three classifiers for $\alpha$:} a kernel support vector regression (kSVR)
\add{from the kernel matrix} with the \textit{kernlab}
package~\autocite{zeileis2004kernlab}, an ordinary SVR \add{from the nLTT
matrix} with the e1071 package~\autocite{meyer2015e1071}, and a linear
regression from the Sackin's index values. The accuracy of each classifier was 
evaluated with 1000 two-fold cross validations \add{with equally-sized folds}.

@@

\del{For each of the four parameters, we also tested a linear regression against
Sackin's index~\autocite{shao1990tree} and an ordinary SVR based on the normalized
lineages-through-time (nLTT) statistic~\autocite{janzen2015approximate}.}

@@

\add{Despite the fact that the parameter values used to generate the simulated
transmission trees were known, the true posterior distributions of the BA
parameters were unknown. Therefore, any apparent errors or biases in the
estimates could be due to either poor performance of our method, or to real
features of the posterior distribution. Two retrospective experiments were
performed to disambiguate some of the observed errors. To assess the impact of
the SMC settings on \textit{netabc}'s accuracy, we ran \textit{netabc}
twice on the same simulated transmission tree. For the first run, the SMC
settings were the same as in the other simulations: 1000 particles, 5 simulated
transmission trees per particle, and $\alpha_{\text{ESS}} = 0.95$. The second
run was performed with 2000 particles, 10 simulated transmission trees per
particle, and $\alpha_{\text{ESS}} = 0.99$. To investigate the extent to which
errors in the estimated BA parameters were due to true features of the
posterior, rather than an inaccurate ABC approximation, we performed marginal
estimation for one set of parameter values. Each combination of 1, 2, or 3
model parameters (14 combinations total) was fixed to their known values, and
the remaining parameters were estimated with \textit{netabc}. The parameter
values were $\alpha = 0.0$, $m = 2$, $I = 2000$, and $N = 5000$.}

@@

\del{Two} \add{Four} of the datasets \del{\autocite{li2015hiv,novitsky2014impact}}
(MSM/Shanghai, HET/Botswana, HET/Uganda, and MSM/USA)
were initially much larger than the others, containing 1265, 1299,
\add{1026/915 (\textit{env}/\textit{gag}), and 648} sequences respectively. To
ensure that the analyses were comparable, we reduced these to a number of
sequences similar to the smaller datasets. For the MSM/Shanghai data, we
detected a cluster of size 280 using a patristic distance cutoff of 0.02 as
described previously~\autocite{poon2015impact}. Only sequences within this
cluster were carried forward. For the HET/Uganda, HET/Botswana, and MSM/USA
data, no large clusters were detected using the same cutoff, so we analysed
\del{a subtree} subsets of sizes 255, 180, and 180 respectively. \add{The
subset of the HET/Uganda data was chosen by eye such that the individuals were
monopolistic in both the \textit{gag} and \textit{env} trees. The other subsets
were arbitrarily chosen subtrees from phylogenies of the complete datasets.}

@@

<<classifiers, include=FALSE>>=
    get.kernel.data <- function (param, step) {
        md <- collect.metadata(paste0('../../simulations/kernel-', param, '/', step, '/*'))
        if ('rbf_variance' %in% colnames(md)) {
          md <- subset(md, rbf_variance == 4 & decay_factor == 0.3)
        }
        if (param == 'alpha') {
            md <- subset(md, m == 2)
        }
        d <- setDT(collect.data(rownames(md)))
        if (param %in% c('alpha', 'm')) {
            d[,list(rsquared=mean(rsquared)), by=c('nsimnode', 'ntip')]
        } else if (param == 'I') {
            d[,list(rsquared=mean(rsquared)), by=ntip]
        } else {
            d[,list(rsquared=mean(rsquared)), by=c('I', 'ntip')]
        }
    }
    kernel.alpha <- get.kernel.data('alpha' ,'classifier')
    kernel.m <- get.kernel.data('m', 'classifier')
    kernel.I <- get.kernel.data('I', 'classifier')
    kernel.N <- get.kernel.data('N', 'classifier')

@@

We investigated the \add{identifiability of four} parameters of the BA network
model~\autocite{barabasi1999emergence}: the number of nodes $N$, the
preferential attachment power $\alpha$, the number of edges added per vertex
$m$, and the number of infected nodes $I$. \del{In addition to $m$ and $\alpha$
(see Introduction), we considered $N$, which denotes the total number of nodes
in the network, and $I$, which is the number of infected nodes at which to stop
the simulation and sample the transmission tree.} To examine the effect of
these parameters on tree shape, we simulated transmission trees under different
parameter values, calculated pairwise tree kernel scores between them, and
attempted to classify the trees using a kernel support vector machine (kSVR).
We also tested classifiers based on Sackin's index~\autocite{shao1990tree} and
the normalized lineages-through-time (nLTT)
statistic~\autocite{janzen2015approximate}. The accuracy of each classifier
\add{on all BA parameters is shown in} \del{varied based on the parameter
being tested} \cref{fig:rsquared}. Classifiers based on \del{two other tree
statistics,} the nLTT and Sackin's index generally exhibited worse performance
than the tree kernel, although the magnitude of the disparity varied between
the parameters (\cref{fig:rsquared}, centre and right). The results were
largely robust to variations in the tree kernel meta-parameters $\lambda$ and
$\sigma$, \add{although accuracy varied between different epidemic and sampling
scenarios} (\cref{fig:alphacrossv,fig:mcrossv,fig:Ncrossv,fig:Icrossv}).
\add{For all parameters except $m$, the absolute number of tips in the tree had
a much greater impact on accuracy than the proportion of infected individuals
these tips represented. However, for $m$, both the number and proportion of
sampled tips had a strong impact or the accuracy of the kSVR
(\cref{fig:mcrossv}).}

\begin{figure}[ht]
    \centering
    \includegraphics[width=\textwidth]{kernel-rsquared.pdf}
    \caption[
        Cross-validation accuracy of kernel-SVR, nLTT-based SVR, and Sackin's
        index regression classifiers for BA model parameters.
    ]{
        Cross-validation accuracy of kernel-SVR classifier (left), SVR
        classifier using nLTT (centre), and linear regression using
        Sackin's index (right) for BA model parameters. Kernel meta-parameters
        were set to $\lambda = 0.3$ and $\sigma = 4$. Each point was calculated
        based on 300 simulated transmission trees over networks with three
        different values of the parameter being tested, assuming perfect
        knowledge of the other parameters. Vertical lines are empirical 95\%
        confidence intervals based on 1000 two-fold cross-validations. \add{The
        classifiers for $I$ were not evaluated with 1000-tip trees, because one
        of the tested $I$ values was 500, and it is not possible to sample a
        tree of size 1000 from 500 infected individuals.}
    }
    \label{fig:rsquared}
\end{figure}

\add{The accuracy of each classifier on all BA parameters is shown in
\cref{fig:rsquared}. Classifiers based on the nLTT and Sackin's index generally
exhibited worse performance than the tree kernel, although the magnitude of the
disparity varied between the parameters (\cref{fig:rsquared}, centre and
right). The results were largely robust to variations in the tree kernel
meta-parameters $\lambda$ and $\sigma$, although accuracy varied between
different epidemic and sampling scenarios
(\cref{fig:alphacrossv,fig:mcrossv,fig:Icrossv,fig:Ncrossv}). For all
parameters except $m$, the absolute number of tips in the tree had a much
greater impact on accuracy than the proportion of infected individuals these
tips represented. However, for $m$, both the number and proportion of sampled
tips had a strong impact or the accuracy of the kSVR (\cref{fig:mcrossv}).}

The kSVR classifier for $\alpha$ had an average $R^2$ of 
    \Sexpr{kernel.alpha[,round(mean(rsquared), 2)]},
compared to 
    \Sexpr{nltt.alpha[,round(mean(rsquared), 2)]}
for the nLTT-based SVR, and
    \Sexpr{sackin.alpha[,round(mean(rsquared), 2)]}
for the linear regression against Sackin's index. There was little variation
about the mean for different tree and epidemic sizes. No classifier could
accurately identify the $m$ parameter in any epidemic scenario, with average
$R^2$ values of 
  \Sexpr{kernel.m[,round(mean(rsquared), 2)]} for kSVR,
  \Sexpr{nltt.m[,round(mean(rsquared), 2)]} for the nLTT, and
  \Sexpr{sackin.m[,round(mean(rsquared), 2)]}
for Sackin's index. Again, there was little variation in accuracy between
epidemic scenarios, although the accuracy of the kSVR was slightly higher
on 1000-tip trees (\cref{fig:rsquared}, left).

>>>>

When an infectious disease spreads through a population, transmissions are
generally more likely to occur between certain pairs of individuals. Such pairs
must have a particular mode of contact with one another, which varies with the
mode of transmission of the disease. For airborne pathogens, physical proximity
may be sufficient, while for sexually transmitted diseases, sexual or in some
cases blood-to-blood contact is required. The population together with the set
of links between individuals along which transmission can occur is called the
contact network~\autocite{klovdahl1985social, morris1993epidemiology}. The
structure of the contact network underlying an epidemic can profoundly impact
the speed and pattern of the epidemic's expansion. Network structure can
influence the prevalence trajectory~\autocite{o2011contact, ma2013importance}
and epidemic threshold~\autocite{barthelemy2005dynamical}, in turn affecting
the estimates of quantities such as effective viral population
size~\autocite{goodreau2006assessing}.  From a public health perspective,
contact networks have been explored as tools for curtailing epidemic spread, by
way of interventions targeted to well-connected
nodes~\autocite{wang2015targeting}. True contact networks are a challenging
type of data to collect, requiring extensive epidemiological
investigation~\autocite{welch2011statistical, eames2015six}.

Viral sequence data, on the other hand, has become \del{relatively inexpensive
and straightforward to collect on a population level.} \add{easier to collect
as the cost of sequencing has declined. In the case of HIV, genotyping has
become part of routine clinical care in several health regions.} Due to the
high mutation rate of RNA viruses, epidemiological processes impact the course
of viral evolution, thereby shaping the inter-host viral
phylogeny~\autocite{drummond2003measurably}. The term ``phylodynamics'' was
coined to describe this interaction, as well as the growing family of inference
methods to estimate epidemiological parameters from viral
phylogenies~\autocite{grenfell2004unifying}. These methods have revealed
diverse properties of local viral outbreaks, from basic reproductive
number~\autocite{stadler2012estimating}, to the degree of
clustering~\autocite{hughes2009molecular}, to the elevated transmission risk
during acute infection~\autocite{volz2012simple}. On the other hand, although
sophisticated methods have been developed for fitting complex population
genetic models to phylogenies~\autocite{volz2012complex,
rasmussen2014phylodynamic}, inference of structural network parameters has to
date been limited. However, it has been shown that network structure has a
tangible impact on phylogeny shape~\autocite{leventhal2012inferring,
colijn2014phylogenetic, goodreau2006assessing, robinson2013dynamics,
villandre2016assessment}, suggesting that such statistical inference might be
possible~\autocite{welch2011statistical}. \add{In the context of networks,
sequence data have the advantage of being objective, in that they are not
affected by misreporting. However, just as with survey data, it is important to
collect a representative sample from the population to perform accurate
inference~\autocite{novitsky2014impact}.}

Survey-based studies of sexual networks~\autocite{colgate1989risk,
liljeros2001web, schneeberger2004scale, latora2006network, rothenberg2007large,
clemenccon2015statistical} have found that these networks tend to have a degree
distribution which follows a power law \autocite[although there has been some
disagreement, see][]{handcock2004likelihood}. That is, the number of nodes of
degree $k$ is proportional to $k^{-\gamma}$ for some constant $\gamma$. These
networks are also referred to as
``scale-free''~\autocite{barabasi1999emergence}. One process by which
scale-free networks can be generated is preferential attachment, where nodes
with a high number of contacts attract new connections at an elevated rate. The
first contact network model incorporating preferential attachment was
introduced by \textcite{barabasi1999emergence}, and is now referred to as the
Barab\'asi-Albert (BA) model. Under this model, networks are formed by
iteratively adding nodes with $m$ new edges each. In the most commonly studied
formulation, these new edges are joined to existing nodes of degree $k$ with
probability proportional to $k$, so that nodes of high degree tend to attract
more connections. \citeauthor{barabasi1999emergence} suggested an extension
where the probability of attaching to a node of degree $k$ is $k^\alpha$ for
some non-negative constant $\alpha$, and we use this extension in this work.
\add{When $\alpha \neq 1$, the degree distribution is no longer a power law:
for $\alpha < 1$, the distribution is a stretched exponential, while for
$\alpha > 1$, it is a ``gelation'' type distribution where one or a few hub
nodes are connected to nearly every other node in the
graph~\autocite{krapivsky2000connectivity}.}

@@

Standard methods of model fitting involve calculation of the likelihood of
observed data under the model. In maximum likelihood estimation, a quantity
proportional to the likelihood is optimized, often through a standard
multi-dimensional numerical optimization procedure. Bayesian methods integrate
prior information by optimizing the posterior probability instead. To avoid
calculation of a normalizing constant, Bayesian inference is often performed
using Markov chain Monte Carlo (MCMC), which uses likelihood \emph{ratios} in
which the normalizing constants cancel out. Unfortunately, it is generally
difficult to explicitly calculate the likelihood of an observed transmission
tree under a contact network model, even up to a normalizing constant. To do
so, it would be necessary to integrate over all possible networks, and also
over all possible labellings of the internal nodes of the transmission tree.
While it is not known (to us) whether such integration is tractable, a simpler
alternative is offered by likelihood-free methods, namely approximate Bayesian
computation (ABC)~\autocite{tavare1997inferring, beaumont2002approximate}. ABC
leverages the fact that, although calculating the likelihood may be
impractical, generating simulated datasets according to a model is often
straightforward. If our model fits the data well, the simulated data it
produces should be similar to the observed data. More formally, if $D$ is the
observed data, the posterior distribution $f(\theta \mid D)$ on model
parameters $\theta$ is replaced as the target of statistical inference by
$f(\theta \mid \rho(\hat{D}, D) < \varepsilon)$, where $\rho$ is a distance
function, $\hat{D}$ is a simulated dataset according to $\theta$, and
$\varepsilon$ is a small tolerance~\autocite{sunnaaker2013approximate}. \del{In
the specific case when $\rho$ is a kernel function, the approach is known as
ABC~\autocite{nakagome2013kernel, poon2015phylodynamic}.} \add{Our
group~\autocite{poon2015phylodynamic} and others~\autocite{park2015k2} have
demonstrated that taking $\rho$ to be a well-chosen kernel function can produce
a more accurate ABC approximation than the typical choice of a difference of
summary statistics.}

@@

\add{In the original formulation of ABC-SMC~\autocite{sisson2007sequential,
beaumont2009adaptive}, the user is required to specify a decreasing sequence of
tolerances $\lbrace\varepsilon_i\rbrace$. At iteration $i$, particles with no
associated simulated datasets within distance $\varepsilon_i$ of the observed
data are removed from the population. In the adaptive version of
\textcite{del2012adaptive}, the sequence of tolerances is determined
automatically by fixing the decay rate of the population's expected sample size
(ESS) to a user-defined value. \citeauthor{del2012adaptive} call this value
$\alpha$, but we will refer to it here as $\alpha_{\text{ESS}}$ to avoid
confusion with the preferential attachment power parameter of the BA model.}

@@

Nodes in our networks followed simple SI dynamics, meaning that they became
infected at a rate proportional to their number of infected neighbours, and
never recovered. For all analyses, the transmission trees' branch lengths were
scaled by dividing by their mean. We used the \textit{igraph} library's
implementation of the BA model~\autocite{csardi2006igraph} to generate the
graphs. The analyses were run on Westgrid (\url{https://www.westgrid.ca/}) and
a local computer cluster. \add{A computer program implementing our method is
freely available at \url{https://github.com/rmcclosk/netabc} (last accessed
\today).}

@@

\add{This experiment also had the secondary purpose of validating our choice of
the tree kernel as a distance measure in ABC. To tell the trees apart, we used
a classifier based on the tree kernel, but we also tested two other tree shape
statistics. Sackin's index~\autocite{shao1990tree} is a measure of tree
imbalance which not take branch lengths into account, considering only the
topology. The normalized lineages-through-time
\autocite[nLTT,][]{janzen2015approximate} compares two trees based on normalized
distributions of their branching times, and does not explicitly consider the
topology. Since the tree kernel incorporates both of these sources of
information, we expected it to outperform the other two statistics. Finally,
the tree kernel can be tuned by adjusting the values of the meta-parameters
$\lambda$ and $\sigma$ (the ``decay factor'' and ``radial basis function
variance'', see \textcite{poon2013mapping}). $\lambda$ is used to penalize
ween large subset trees which tend to dominate the kernel score. When $\lambda
= 0$, all but the root branches of each subset tree are ignored, while when
$\lambda = 1$, no penalty is applied. $\sigma$ controls how strictly the notion
of similarity is applied to branch lengths. When $\sigma = 0$, branch lengths
must match exactly, while as $\sigma \to \infty$, branch lengths are not
considered at all.}

\del{We used the phylogenetic kernel developed by \textcite{poon2013mapping}}
To test whether the parameters of the BA model had a \add{measurable} effect on
tree shape, 100 networks were simulated under each of three different values of
$\alpha$: 0.5, 1.0, and 1.5 (300 networks total). The other parameters were
fixed to the following values: $N = 5000$, $I = 1000$, and $m = 2$. A
transmission tree with 500 tips was simulated over each network (300
transmission trees total). The 300 trees were compared pairwise with the tree
kernel to form a $300 \times 300$ kernel matrix. The kernel meta-parameters
$\lambda$ \del{(the ``decay factor''),} and $\sigma$ \del{(the ``radial basis
function variance'')~\autocite[see][]{poon2013mapping},} were set to 0.3 and 4
respectively. \add{We also computed a $300 \times 300$ matrix of pairwise nLTT
values, and a $1 \times 300$ vector of Sackin's index values.} We constructed 
\add{three classifiers for $\alpha$:} a kernel support vector regression (kSVM)
\add{from the kernel matrix} with the \textit{kernlab}
package~\autocite{zeileis2004kernlab}, and two ordinary SVMs \add{from the nLTT
matrix and Sackin's index vector} with the e1071
package~\autocite{meyer2015e1071}.\del{, and a linear regression from the
Sackin's index values.} The accuracy of each classifier was evaluated with 1000
two-fold cross validations \add{with equally-sized folds}.

@@

\del{For each of the four parameters, we also tested a linear regression against
Sackin's index~\autocite{shao1990tree} and an ordinary SVM based on the normalized
lineages-through-time (nLTT) statistic~\autocite{janzen2015approximate}.}

@@

\add{Despite the fact that the parameter values used to generate the simulated
transmission trees were known, the true posterior distributions of the BA
parameters were unknown. Therefore, any apparent errors or biases in the
estimates could be due to either poor performance of our method, or to real
features of the posterior distribution. Two retrospective experiments were
performed to disambiguate some of the observed errors. To assess the impact of
the SMC settings on \textit{netabc}'s accuracy, we ran \textit{netabc}
twice on the same simulated transmission tree. For the first run, the SMC
settings were the same as in the other simulations: 1000 particles, 5 simulated
transmission trees per particle, and $\alpha_{\text{ESS}} = 0.95$. The second
run was performed with 2000 particles, 10 simulated transmission trees per
particle, and $\alpha_{\text{ESS}} = 0.97$. To investigate the extent to which
errors in the estimated BA parameters were due to true features of the
posterior, rather than an inaccurate ABC approximation, we performed marginal
estimation for one set of parameter values. Each combination of 1, 2, or 3
model parameters (14 combinations total) was fixed to their known values, and
the remaining parameters were estimated with \textit{netabc}. The parameter
values were $\alpha = 0.0$, $m = 2$, $I = 2000$, and $N = 5000$.}

@@

\del{Two} \add{Four} of the datasets \del{\autocite{li2015hiv,novitsky2014impact}}
(MSM/Shanghai, HET/Botswana, HET/Uganda, and MSM/USA)
were initially much larger than the others, containing 1265, 1299,
\add{1026/915 (\textit{env}/\textit{gag}), and 648} sequences respectively. To
ensure that the analyses were comparable, we reduced these to a number of
sequences similar to the smaller datasets. For the MSM/Shanghai data, we
detected a cluster of size 280 using a patristic distance cutoff of 0.02 as
described previously~\autocite{poon2015impact}. Only sequences within this
cluster were carried forward. For the HET/Uganda, HET/Botswana, and MSM/USA
data, no large clusters were detected using the same cutoff, so we analysed
\del{a subtree} subsets of sizes 255, 180, and 180 respectively. \add{The
subset of the HET/Uganda data was chosen by eye such that the individuals were
monophyletic in both the \textit{gag} and \textit{env} trees. The other
subsets were arbitrarily chosen subtrees from phylogenies of the complete
datasets.}

@@

<<classifiers, include=FALSE>>=
    get.kernel.data <- function (param, step) {
        md <- collect.metadata(paste0('../../simulations/kernel-', param, '/', step, '/*'))
        if ('rbf_variance' %in% colnames(md)) {
          md <- subset(md, rbf_variance == 4 & decay_factor == 0.3)
        }
        if (param == 'alpha') {
            md <- subset(md, m == 2)
        }
        d <- setDT(collect.data(rownames(md)))
        if (param == 'alpha') {
            d[,list(rsquared=mean(rsquared), accuracy=mean(accuracy)), by=c('nsimnode', 'ntip')]
        } else if (param == 'm') {
            d[,list(rsquared=mean(rsquared)), by=c('nsimnode', 'ntip')]
        } else if (param == 'I') {
            d[,list(rsquared=mean(rsquared), accuracy=mean(accuracy)), by=ntip]
        } else {
            d[,list(rsquared=mean(rsquared)), by=c('I', 'ntip')]
        }
    }
    kernel.alpha <- get.kernel.data('alpha' ,'classifier')
    kernel.m <- get.kernel.data('m', 'classifier')
    kernel.I <- get.kernel.data('I', 'classifier')
    kernel.N <- get.kernel.data('N', 'classifier')

@@

We investigated the \add{identifiability of four} parameters of the BA network
model~\autocite{barabasi1999emergence}: the number of nodes $N$, the
preferential attachment power $\alpha$, the number of edges added per vertex
$m$, and the number of infected nodes $I$. \del{In addition to $m$ and $\alpha$
(see Introduction), we considered $N$, which denotes the total number of nodes
in the network, and $I$, which is the number of infected nodes at which to stop
the simulation and sample the transmission tree.} To examine the effect of
these parameters on tree shape, we simulated transmission trees under different
parameter values, calculated pairwise tree kernel scores between them, and
attempted to classify the trees using a kernel support vector machine (kSVM).
We also tested classifiers based on Sackin's index~\autocite{shao1990tree} and
the normalized lineages-through-time (nLTT)
statistic~\autocite{janzen2015approximate}. \add{We report the accuracy of the
classifiers, which is simply the proportion of trees which were assigned the
correct parameter value. Since there were three possible values, random
guessing would produce an accuracy of 0.33.} \del{The accuracy of each
classifier varied based on the parameter being tested} \add{The results are
shown in} \cref{fig:rsquared}. Classifiers based on \del{two other tree
statistics,} the nLTT and Sackin's index generally exhibited worse performance
than the tree kernel, although the magnitude of the disparity varied between
the parameters (\cref{fig:rsquared}, centre and right). \del{The results were
largely robust to variations in the tree kernel meta-parameters $\lambda$ and
$\sigma$, although accuracy varied between different epidemic and sampling
scenarios} \add{Larger datasets were generally classified more accurately
(\cref{fig:alphacrossv,fig:mcrossv,fig:Ncrossv,fig:Icrossv}), although large
values of $\lambda$ produced worse estimates on large datasets. Extremely low
$\sigma$ values, which require nearly-exact matches between branch lengths,
resulted in low accuracy in some cases (\textit{e.g.} \cref{fig:alphacrossvn},
center row).}

\begin{figure}[ht]
    \centering
    \includegraphics[width=\textwidth]{kernel-rsquared.pdf}
    \caption[
        Cross-validation accuracy of kernel-SVM, nLTT-based SVM, and Sackin's
        index regression classifiers for BA model parameters.
    ]{
        Cross-validation accuracy of kernel-SVM classifier (left), SVM
        classifiers using nLTT (centre) and Sackin's index (right) for BA model
        parameters. Kernel meta-parameters were set to $\lambda = 0.3$ and
        $\sigma = 4$. Each point was calculated based on 300 simulated
        transmission trees over networks with three different values of the
        parameter being tested, assuming perfect knowledge of the other
        parameters. Vertical lines are empirical 95\% confidence intervals
        based on 1000 two-fold cross-validations. \add{The classifiers for $I$
        were not evaluated with 1000-tip trees, because one of the tested $I$
        values was 500, and it is not possible to sample a tree of size 1000
        from 500 infected individuals.}
    }
    \label{fig:rsquared}
\end{figure}

The kSVM classifier for $\alpha$ had an average accuracy of 
    \Sexpr{kernel.alpha[,round(mean(accuracy), 2)]},
compared to 
    \Sexpr{nltt.alpha[,round(mean(accuracy), 2)]}
for the nLTT, and
    \Sexpr{sackin.alpha[,round(mean(accuracy), 2)]}
for Sackin's index. \del{There was little variation about the mean for
different tree and epidemic sizes.} No classifier could accurately identify $m$
in any epidemic scenario, with average $R^2$ values of 
  \Sexpr{kernel.m[,round(mean(rsquared), 2)]} for kSVM,
  \Sexpr{nltt.m[,round(mean(rsquared), 2)]} for the nLTT, and
  \Sexpr{sackin.m[,round(mean(rsquared), 2)]}
for Sackin's index. There was little variation in accuracy between epidemic
scenarios, although the accuracy of the kSVM was slightly higher on 1000-tip
trees (\cref{fig:rsquared}, left).

The accuracy of classifiers for $I$ varied significantly with the number of
tips in the tree. For 100-tip trees, the average accuracy was
  \Sexpr{kernel.I[ntip == 100, round(mean(accuracy), 2)]},
  \Sexpr{nltt.I[ntip == 100, round(mean(accuracy), 2)]}, and
  \Sexpr{sackin.I[ntip == 100, round(mean(accuracy), 2)]}
for the tree kernel, nLTT, and Sackin's index respectively.
For 500-tip
trees, the values increased to
  \Sexpr{kernel.I[ntip == 500, round(mean(accuracy), 2)]},
  \Sexpr{nltt.I[ntip == 500, round(mean(accuracy), 2)]}, and
  \Sexpr{sackin.I[ntip == 500, round(mean(accuracy), 2)]}.
Finally, the performance of classifiers for $N$ depended heavily on the
epidemic scenario. The $R^2$ of the kSVM classifier ranged from
  \Sexpr{kernel.N[,round(min(rsquared), 2)]}
for the smallest epidemic and smallest sample size, to
  \Sexpr{kernel.N[,round(max(rsquared), 2)]}
for the largest. Likewise, $R^2$ for the nLTT-based SVM ranged from 
  \Sexpr{nltt.N[,round(min(rsquared), 2)]}
to
  \Sexpr{nltt.N[,round(max(rsquared), 2)]}.
Sackin's index did not accurately classify $N$ in any scenario, with an average
$R^2$ of
  \Sexpr{sackin.N[,round(mean(rsquared), 2)]}
and little variation between scenarios.
